Finch: Sparse and Structured Array Programming with Control Flow

### 2024-03-22 sections/eval.tex
Update on Overleaf.

In this case study, we highlight a few different Finch formats, as specified in Table \ref{spmv_tensor_formats}, and the performance effects of conforming a dataset’s structure with its storage format, which Finch's datastructure-driven model enables us to do. Finch also provides the control flow necessary to manipulate data reads and writes, enabling exploitation of multiple structural patterns concurrently (e.g. sparsity \textit{and} symmetry).

@@

\subsubsection{Tensor Formats}
We found that the SpMV performance was superior for the level format that best paralleled the structure of the tensor. Namely, matrices with a clear blocked structure like exdata\_1, TSOPF\_RS\_b678\_c1, and heart3 performed notably well with the SparseVBL format in comparison to both TACO and the Finch SparseList formats (which we consider the closest analog to the sparse matrix format used in TACO and other libraries) with speedups of 2.16, 1.55, and 1.30 relative to TACO and of 1.47, 2.13, and 1.42 relative to the fastest SparseList Finch SpMV implementation. Furthermore, the synthetic Toeplitz banded matrices we constructed performed the best with the SparseBand matrix, in particular with the toeplitz\_large\_band and the toeplitz\_medium\_band matrices having a speedup of 1.98 and 1.64 relative to TACO and a speedup of 1.47 and 1.24 relative to the fastest SparseList Finch SpMV implementation.

There were also significant advantages of using the Pattern format instead of the Element format to represent scalar values in the matrices when these values were boolean, such as matrices in the SNAP collection which represent graph datasets are boolean. For example, Finch Pattern SpMV had a speedup of 2.51 over TACO for SNAP/email-Eu-core and a speedup of X over the same Finch program, but with an Element format.

\begin{table}[htbp]
    \centering
    \caption{SpMV Tensor Formats}
    \label{spmv_tensor_formats}
    \begin{tabular}{|l|l|l|}
        \hline
        \textbf{Spy} & \textbf{Group / Name} & \textbf{Dimensions}\\
        \hline
        & HB/saylr4 & 3,564 x 3,564 (22316) \\
        \hline 
    \end{tabular}
\end{table}

\subsubsection{Symmetric SpMV}
Finch enables us to exploit symmetry in the sparse matrix of the SpMV kernel by providing the capabilities to reuse memory reads and insert control flow logic to restrict iterations to either the lower or upper triangle of the sparse matrix. We can apply this strategy with any level format. Every symmetric matrix in the SparseList and SparseList-Pattern formats has better performance when we use a Finch SpMV program that takes advantage of this symmetry. However, the regular Finch SpMV program has better performance for symmetric matrices than the symmetric Finch SpMV program for the other more specialized formats, likely because we need in-order accesses to fully capitalize on the specialized storage. Symmetric SpMV with the SparseList level format in Finch results in an average of 1.30x speedup over TACO and symmetric SpMV with the SparseList-Pattern format in Finch results in an average speedup of 1.15x over TACO. Notably, there is a 1.90x speedup for the HB/saylr4 matrix over TACO.

@@

\begin{figure}
    \includegraphics[width=\linewidth]{spmv_performance_grouped.png}
    \caption{Performance of SpMV by Finch format.}
    \label{spmv_grouped}
    \footnotesize The performance displayed for Finch on each dataset is the fastest among the formats we tested. "R" indicates row-major implementation and "C" indicates column-major implementation in Finch.
\end{figure}

>>>>

In this case study, we highlight a few different Finch formats as specified in Table \ref{spmv_tensor_formats}, and the performance effects of conforming a dataset’s structure with its storage format, which Finch's datastructure-driven model enables us to do. Finch also provides the control flow necessary to manipulate data reads and writes, enabling exploitation of multiple structural patterns concurrently (e.g. sparsity \textit{and} symmetry).

@@

\subsubsection{Tensor Formats}
We found that the SpMV performance was superior for the level format that best paralleled the structure of the tensor. We consider the Dense(SparseList(Element)) format with the column-major SpMV program to be the Finch baseline as it is the closet analog to the sparse matrix format and SpMV program in other libraries.

Namely, matrices with a clear blocked structure like exdata\_1, TSOPF\_RS\_b678\_c1, and heart3 performed notably well with the SparseVBL format with speedups of 2.16, 1.55, and 1.30 relative to TACO, while the baseline format had slowdowns of 0.71, 0.53, and 0.92 relative to TACO. Furthermore, the synthetic Toeplitz banded matrices we constructed performed the best with the SparseBand matrix, in particular with the toeplitz\_large\_band and the toeplitz\_medium\_band matrices having a speedup of 1.98 and 1.64 relative to TACO, while the baseline format had slowdowns of 0.84 and 0.51 relative to TACO.

There were also significant advantages of using the Pattern format instead of the Element format to represent scalar values in the matrices when these values were boolean, such as matrices in the SNAP collection which represent graph datasets are boolean. For example, the SparseList-Pattern for email-Eu-core resulted in a speedup of 2.51, while the SparseList-Element format resulted in a slowdown of 1.84 over TACO.

\begin{table}[htbp]
    \centering
    \caption{SpMV Sample Matrices}
    \label{spmv_sample_matrices}
    \begin{tabular}{|m{3cm}|c|c|c|}
        \hline
        \textbf{Spy} & \textbf{Group / Name} & \textbf{Dimensions} & \textbf{Best Finch Format}\\
        \hline
        \includegraphics[width=3cm]{spmv_matrices/saylr4.png} & HB/saylr4 & 3,564 x 3,564 (22,316) & Symmetric SparseList \\
        \hline 
        \includegraphics[width=3cm]{spmv_matrices/heart3.png} & Norris/heart3 & 2,339 x 2,339 (680,341) & SparseVBL \\
        \hline 
        \includegraphics[width=3cm]{spmv_matrices/as-735.png} & SNAP/as-735 & 7,716 x 7,716 (26,467) & Symmetric SparseList-Pattern \\
        \hline
    \end{tabular}
\end{table}

\subsubsection{Symmetric SpMV}
Finch enables us to exploit symmetry in the sparse matrix of the SpMV kernel by providing the capabilities to reuse memory reads and insert control flow logic to restrict iterations to either the lower or upper triangle of the sparse matrix. We can apply this strategy with any level format. Every symmetric matrix in the SparseList and SparseList-Pattern formats has better performance when we use a Finch SpMV program that takes advantage of this symmetry. However, the regular row- or column-major Finch SpMV programs have better performance for symmetric matrices than the symmetric Finch SpMV program for the other more specialized formats, likely because we need in-order accesses to fully capitalize on the specialized storage. Symmetric SpMV with the SparseList level format in Finch results in an average of 1.27x speedup over TACO and symmetric SpMV with the SparseList-Pattern format in Finch results in an average speedup of 1.21x over TACO. Notably, there is a 1.91x speedup for the HB/saylr4 matrix over TACO.

@@

\begin{figure}
    \includegraphics[width=\linewidth]{spmv_performance_grouped.png}
    \caption{Performance of SpMV by Finch format.}
    \label{spmv_grouped}
    \footnotesize The performance displayed for Finch on each dataset is the fastest among the formats we tested. "R" indicates row-major implementation and "C" indicates column-major implementation in Finch. The baseline Finch format is unsymmetric Dense(SparseList(Element)).
\end{figure}
